---
title: "Fig1E"
output: html_document
date: "2024-07-08"
---

read in the metadata and all the csv's in /data_7_8_24
note that the protein dataframes have different columns (one has PG.uniprotIDs and the other has PG.ProteinAccessions)
```{r}


# Load the library
library(clusterProfiler)
library(org.Hs.eg.db)

# Verify installation
packageVersion("clusterProfiler")

```


```{r}

library(readr)
Mitra_Fig1_controls_precursor_Report_SN19 <- read_csv("20240702_231826_6010_Mitra_Fig1_controls_precursor_Report_SN19.csv")

Mitra_Fig1_controls_protein_Report_SN19 <- read_csv("20240702_231826_6010_Mitra_Fig1_controls_protein_Report_SN19.csv")

plasma_Fig1_controls_precursor_Report_SN19 <- read_csv("20240702_232929_6010_plasma_Fig1_controls_precursor_Report_SN19.csv")

plasma_Fig1_controls_protein_Report_SN19 <- read_csv("20240702_232929_6010_plasma_Fig1_controls_protein_Report_SN19.csv")
ls
metadata_Mitra_proteomics <- read_csv("metadata_Mitra_proteomics.csv")

# rename the column uniprotID's to ProteinAccessions so the column names match
### THIS SHOULD BE FIXED AND DATA RE-EXPORTED AND RE-ANALYZED
colnames(Mitra_Fig1_controls_protein_Report_SN19)[colnames(Mitra_Fig1_controls_protein_Report_SN19) == "PG.UniProtIds"] <- "PG.ProteinAccessions"


```


Read in the human swissprot transmembrane and subcellular locations proteome
```{r}

swissprot_transmembrane_human <- read_csv("swissprot_transmembrane_human.csv")
swissprot_subcellular_location_human <- read_csv("swissprot_subcellular_location_human.csv")

```


The uniprot accession column in the Mitra/plasma dataframes has some rows with more than one accession
these are separated by a semicolon
they appear redundant, so we will keep only the string before the semicolon

```{r}

library(dplyr)

# In column PG.ProteinAccessions, keep only the first string (remove redundant entries for the uniprot accessions)
Mitra_Fig1_controls_protein_Report_SN19 <- Mitra_Fig1_controls_protein_Report_SN19 %>%
  mutate(PG.ProteinAccessions = sapply(strsplit(PG.ProteinAccessions, ";"), `[`, 1))

plasma_Fig1_controls_protein_Report_SN19 <- plasma_Fig1_controls_protein_Report_SN19 %>%
  mutate(PG.ProteinAccessions = sapply(strsplit(PG.ProteinAccessions, ";"), `[`, 1))

```



Determine number of transmembrane proteins present in the Mitra vs plasma
```{r}

# Initialize the Transmembrane column with "No"
Mitra_Fig1_controls_protein_Report_SN19$Transmembrane <- "No"

# Update Transmembrane column to "Yes" where there is a match
Mitra_Fig1_controls_protein_Report_SN19$Transmembrane[Mitra_Fig1_controls_protein_Report_SN19$PG.ProteinAccessions %in% swissprot_transmembrane_human$Entry] <- "Yes"

# Initialize the Transmembrane column with "No"
plasma_Fig1_controls_protein_Report_SN19$Transmembrane <- "No"

# Update Transmembrane column to "Yes" where there is a match
plasma_Fig1_controls_protein_Report_SN19$Transmembrane[plasma_Fig1_controls_protein_Report_SN19$PG.ProteinAccessions %in% swissprot_transmembrane_human$Entry] <- "Yes"

# Visualize by barplot

# Load ggplot2 for plotting
library(ggplot2)

# Create a barplot for Mitra_Fig1_controls_protein_Report_SN19
ggplot(Mitra_Fig1_controls_protein_Report_SN19, aes(x = Transmembrane)) +
  geom_bar(fill = "steelblue") +
  ggtitle("Transmembrane Status in Mitra_Fig1_controls_protein_Report_SN19") +
  xlab("Transmembrane") +
  ylab("Count")

# Create a barplot for plasma_Fig1_controls_protein_Report_SN19
ggplot(plasma_Fig1_controls_protein_Report_SN19, aes(x = Transmembrane)) +
  geom_bar(fill = "steelblue") +
  ggtitle("Transmembrane Status in plasma_Fig1_controls_protein_Report_SN19") +
  xlab("Transmembrane") +
  ylab("Count")

```

Bring subcellular location data over to the Mitra/plasma protein dataframes
```{r}

# Load the necessary libraries
library(dplyr)
library(tidyverse)

# Perform the left join to add the "Subcellular location [CC]" information
Mitra_Fig1_controls_protein_Report_SN19 <- Mitra_Fig1_controls_protein_Report_SN19 %>%
  dplyr::left_join(
    swissprot_subcellular_location_human %>% 
      dplyr::select(Entry, `Subcellular location [CC]`),
    by = c("PG.ProteinAccessions" = "Entry")
  )


plasma_Fig1_controls_protein_Report_SN19 <- plasma_Fig1_controls_protein_Report_SN19 %>%
  dplyr::left_join(
    swissprot_subcellular_location_human %>% 
      dplyr::select(Entry, `Subcellular location [CC]`),
    by = c("PG.ProteinAccessions" = "Entry")
  )


```


examine structure of the subcellular location column so it can be simplified
```{r}

# Print the first 200 rows of the Subcellular location [CC] column
head(Mitra_Fig1_controls_protein_Report_SN19$`Subcellular location [CC]`, 100)

```


simplify the subcellular location column terms
```{r}

# Load necessary library
library(stringr)

# Updated list of common subcellular locations
locations <- c("Secreted", "Cytoplasm", "Nucleus", "Golgi", "Endoplasmic", "Mitochondrion", 
               "Lysosome", "Peroxisome", "Chromosome", "Endosome", "Cytoskeleton", "Vesicle",
               "Lipid droplet", "Microsome", "Ciliary axoneme", "Cytoplasmic vesicle",
               "Extracellular space", "Peripheral membrane", "Single-pass membrane",
               "Multi-pass membrane", "Membrane raft", "Cell junction", "Caveola", "Focal adhesion",
               "Cilium", "Podosome", "Centrosome", "Spindle")

# Function to simplify subcellular location information
simplify_location <- function(location_string) {
  if (is.na(location_string)) return(NA)
  
  # Extract main locations using regular expressions
  extracted_locations <- str_extract_all(location_string, paste(locations, collapse = "|"))[[1]]
  
  # If no locations found, return NA
  if (length(extracted_locations) == 0) return(NA)
  
  # Create a simplified string preserving hierarchical information
  unique_locations <- unique(extracted_locations)
  simplified_string <- paste(unique_locations, collapse = ", ")
  
  return(simplified_string)
}

# Apply the function to the Subcellular location [CC] column
# Mitra
Mitra_Fig1_controls_protein_Report_SN19$Simplified_Subcellular_Location <- sapply(Mitra_Fig1_controls_protein_Report_SN19$`Subcellular location [CC]`, simplify_location)

# Apply the function to the Subcellular location [CC] column
# plasma
plasma_Fig1_controls_protein_Report_SN19$Simplified_Subcellular_Location <- sapply(plasma_Fig1_controls_protein_Report_SN19$`Subcellular location [CC]`, simplify_location)

# Show the frequency table for the Simplified_Subcellular_Location column, including NAs
# Mitra
table(Mitra_Fig1_controls_protein_Report_SN19$Simplified_Subcellular_Location, useNA = "ifany")

# Show the frequency table for the Simplified_Subcellular_Location column, including NAs
# plasma
table(plasma_Fig1_controls_protein_Report_SN19$Simplified_Subcellular_Location, useNA = "ifany")

# Function to split the Simplified_Subcellular_Location column at each semicolon and duplicate rows
split_and_duplicate <- function(df) {
  df %>%
    mutate(Simplified_Subcellular_Location = ifelse(is.na(Simplified_Subcellular_Location), "NA", Simplified_Subcellular_Location)) %>%
    separate_rows(Simplified_Subcellular_Location, sep = ", ") %>%
    mutate(Simplified_Subcellular_Location = ifelse(Simplified_Subcellular_Location == "NA", NA, Simplified_Subcellular_Location))
}

# Apply the function to the dataframe
# Mitra
Mitra_Fig1_controls_protein_Report_SN19 <- split_and_duplicate(Mitra_Fig1_controls_protein_Report_SN19)
# plasma
plasma_Fig1_controls_protein_Report_SN19 <- split_and_duplicate(plasma_Fig1_controls_protein_Report_SN19)

# Show the frequency table for the Simplified_Subcellular_Location column, including NAs
# this is after the columns are split up
# Mitra
table(Mitra_Fig1_controls_protein_Report_SN19$Simplified_Subcellular_Location, useNA = "ifany")
# plasma
table(plasma_Fig1_controls_protein_Report_SN19$Simplified_Subcellular_Location, useNA = "ifany")

```


Visualize the differences in subcellular representation between plasma and mitra
```{r}

# Load necessary libraries
library(dplyr)
library(ggplot2)
library(tidyr)

# Combine the two dataframes with an additional column to indicate the source
mitra_df <- Mitra_Fig1_controls_protein_Report_SN19 %>%
  mutate(Source = "Blood")

plasma_df <- plasma_Fig1_controls_protein_Report_SN19 %>%
  mutate(Source = "Plasma")

combined_df <- bind_rows(mitra_df, plasma_df)

# Create a frequency table for the Simplified_Subcellular_Location column in each dataframe
freq_table <- combined_df %>%
  group_by(Source, Simplified_Subcellular_Location) %>%
  summarise(Frequency = n()) %>%
  ungroup()

# Replace NA with "NA" for plotting
freq_table$Simplified_Subcellular_Location <- ifelse(is.na(freq_table$Simplified_Subcellular_Location), "Other", freq_table$Simplified_Subcellular_Location)

# Create the bar plot
p <- ggplot(freq_table, aes(x = Simplified_Subcellular_Location, y = Frequency, fill = Source)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, face = "bold")) +
  labs(y = "Frequency", x = NULL, fill = "Source") +
  scale_y_continuous(breaks = seq(0, max(freq_table$Frequency), by = 100)) +
  theme(legend.position = c(0.4, 0.6),
        legend.title = element_text(size = 8),  # Smaller legend title
        legend.text = element_text(size = 8),  # Smaller legend text
        legend.key.size = unit(0.4, "cm"))  # Smaller legend keys


ggsave("subcellular_locations_plot.png", plot = p, width = 5, height = 3, dpi = 600)

rm(plasma_df)
rm(mitra_df)

```

refine the gene name list so it's just one gene for each
Remove any rows with NA for PG.Gene
```{r}

library(dplyr)

# Function to keep only the first gene before a semicolon
clean_genes <- function(gene_column) {
  sapply(gene_column, function(x) strsplit(as.character(x), ";")[[1]][1])
}

# Apply the function to the PG.Genes column in both dataframes
plasma_Fig1_controls_protein_Report_SN19$PG.Genes <- clean_genes(plasma_Fig1_controls_protein_Report_SN19$PG.Genes)
Mitra_Fig1_controls_protein_Report_SN19$PG.Genes <- clean_genes(Mitra_Fig1_controls_protein_Report_SN19$PG.Genes)

# Prepare the lists of genes/proteins for Plasma and Mitra samples, removing NA values
plasma_genes <- unique(na.omit(plasma_Fig1_controls_protein_Report_SN19$PG.Genes))
mitra_genes <- unique(na.omit(Mitra_Fig1_controls_protein_Report_SN19$PG.Genes))

# Inspect the first few entries of the gene lists
cat("First few entries of Plasma Genes:\n")
print(head(plasma_genes, 20))

cat("\nFirst few entries of Mitra Genes:\n")
print(head(mitra_genes, 20))

```

ORA
```{r}

# Load necessary libraries
library(clusterProfiler)
library(org.Hs.eg.db)
library(dplyr)

# Prepare the lists of genes/proteins for Plasma and Mitra samples, removing NA values
plasma_genes <- unique(na.omit(plasma_Fig1_controls_protein_Report_SN19$PG.Genes))
mitra_genes <- unique(na.omit(Mitra_Fig1_controls_protein_Report_SN19$PG.Genes))

# Ensure gene names are in uppercase (for consistency with the GO database)
plasma_genes <- toupper(plasma_genes)
mitra_genes <- toupper(mitra_genes)

# Perform ORA for Plasma sample using GO terms
ora_plasma_go <- enrichGO(gene = plasma_genes,
                          OrgDb = org.Hs.eg.db,
                          keyType = "SYMBOL",
                          ont = "ALL",
                          pvalueCutoff = 0.05,
                          qvalueCutoff = 0.05)

# Perform ORA for Mitra sample using GO terms
ora_mitra_go <- enrichGO(gene = mitra_genes,
                         OrgDb = org.Hs.eg.db,
                         keyType = "SYMBOL",
                         ont = "ALL",
                         pvalueCutoff = 0.05,
                         qvalueCutoff = 0.05)

# View results
cat("ORA Results for Plasma Sample (GO Terms):\n")
print(head(ora_plasma_go))

cat("\nORA Results for Mitra Sample (GO Terms):\n")
print(head(ora_mitra_go))

# Plot results
barplot(ora_plasma_go, showCategory = 20, title = "ORA for Plasma Sample (GO Terms)")
barplot(ora_mitra_go, showCategory = 20, title = "ORA for Mitra Sample (GO Terms)")


```

dotplot
```{r}

# Generate dot plot for Plasma sample ORA results
dotplot(ora_plasma_go, showCategory = 20, title = "ORA for Plasma Sample (GO Terms)")

# Generate dot plot for Mitra sample ORA results
dotplot(ora_mitra_go, showCategory = 20, title = "ORA for Mitra Sample (GO Terms)")

```


CC, BP, MF separately
```{r}


# Perform ORA for Plasma sample using GO terms for CC, MF, and BP separately
ora_plasma_go_cc <- enrichGO(gene = plasma_genes,
                             OrgDb = org.Hs.eg.db,
                             keyType = "SYMBOL",
                             ont = "CC",
                             pvalueCutoff = 0.05,
                             qvalueCutoff = 0.05)

ora_plasma_go_mf <- enrichGO(gene = plasma_genes,
                             OrgDb = org.Hs.eg.db,
                             keyType = "SYMBOL",
                             ont = "MF",
                             pvalueCutoff = 0.05,
                             qvalueCutoff = 0.05)

ora_plasma_go_bp <- enrichGO(gene = plasma_genes,
                             OrgDb = org.Hs.eg.db,
                             keyType = "SYMBOL",
                             ont = "BP",
                             pvalueCutoff = 0.05,
                             qvalueCutoff = 0.05)

# Perform ORA for Mitra sample using GO terms for CC, MF, and BP separately
ora_mitra_go_cc <- enrichGO(gene = mitra_genes,
                            OrgDb = org.Hs.eg.db,
                            keyType = "SYMBOL",
                            ont = "CC",
                            pvalueCutoff = 0.05,
                            qvalueCutoff = 0.05)

ora_mitra_go_mf <- enrichGO(gene = mitra_genes,
                            OrgDb = org.Hs.eg.db,
                            keyType = "SYMBOL",
                            ont = "MF",
                            pvalueCutoff = 0.05,
                            qvalueCutoff = 0.05)

ora_mitra_go_bp <- enrichGO(gene = mitra_genes,
                            OrgDb = org.Hs.eg.db,
                            keyType = "SYMBOL",
                            ont = "BP",
                            pvalueCutoff = 0.05,
                            qvalueCutoff = 0.05)

# View results for Plasma sample
cat("ORA Results for Plasma Sample (GO CC Terms):\n")
print(head(ora_plasma_go_cc))

cat("\nORA Results for Plasma Sample (GO MF Terms):\n")
print(head(ora_plasma_go_mf))

cat("\nORA Results for Plasma Sample (GO BP Terms):\n")
print(head(ora_plasma_go_bp))

# View results for Mitra sample
cat("ORA Results for Mitra Sample (GO CC Terms):\n")
print(head(ora_mitra_go_cc))

cat("\nORA Results for Mitra Sample (GO MF Terms):\n")
print(head(ora_mitra_go_mf))

cat("\nORA Results for Mitra Sample (GO BP Terms):\n")
print(head(ora_mitra_go_bp))

# Plot results as dot plots
dotplot(ora_plasma_go_cc, showCategory = 20, title = "ORA for Plasma Sample (GO CC Terms)", 
        color = "p.adjust", size = "GeneRatio")

dotplot(ora_plasma_go_mf, showCategory = 20, title = "ORA for Plasma Sample (GO MF Terms)", 
        color = "p.adjust", size = "GeneRatio")

dotplot(ora_plasma_go_bp, showCategory = 20, title = "ORA for Plasma Sample (GO BP Terms)", 
        color = "p.adjust", size = "GeneRatio")

dotplot(ora_mitra_go_cc, showCategory = 20, title = "ORA for Mitra Sample (GO CC Terms)", 
        color = "p.adjust", size = "GeneRatio")

dotplot(ora_mitra_go_mf, showCategory = 20, title = "ORA for Mitra Sample (GO MF Terms)", 
        color = "p.adjust", size = "GeneRatio")

dotplot(ora_mitra_go_bp, showCategory = 20, title = "ORA for Mitra Sample (GO BP Terms)", 
        color = "p.adjust", size = "GeneRatio")


```


